Practical Optional Types for Clojure

### 2015-05-15 intro.tex
corrections
delete ns form. delete some type hints. use `new` over dot.

\begin{figure}
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/parent2.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

Figure~\ref{fig:ex1} presents a simple program demonstrating many
aspects of our system, from simple type annotations to explicit
handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
extended form of occurrence typing and Clojure's \emph{type hints},
which are central to Typed Clojure's approach to interoperability.

The \clj{parent} function has the type 
$$
\clj{['{:file (U nil File)} -> (U nil Str)]}
$$
which means that it takes a hash table whose \clj{:file} key maps to either
\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
as an accessor to get the file, checks that it isn't \clj{nil}, and
then obtains the parent by making a Java method call.
%
The annotation \clj{^File f} is a type hint on \clj{f}, which instructs the Clojure
compiler (running prior to Typed Clojure typechecking) to statically
resolve the \clj{getParent} 
call to \clj{File}'s \clj{getParent} method with signature \java{String getParent();}, rather than using reflection at runtime.

@@

%$$
%\abs{\x{}}{\Union{\Nil}{\File}}
%{\letexp {\hinted{\File{}}{f}} {f}
%  {\ifexp {f} 
%    {\letexp {s} {\methodstaticexp {\File{}} {\String{}} {\String{}} {getParent} {f}}
%      {\doexp {\ifexp {s}
%                {\nil{}}
%                {\errorval{input}}}
%              {\newstaticexp {\String{}} {\File{}} {\File{}} {s}}}}
%    {\nil{}}
%}}
%              $$

%(some-> f (.getParent) (File.) (.getParent))

%\begin{lstlisting}
%(ann parent [(U nil File) -> (U nil String)])
%(defn parent [^File f]
%  (when f
%    (.getParent f)))
%\end{lstlisting}

%% \begin{minted}[]{clojure}
%% (defalias Expr
%%   (U '{:op :if :test Expr :then Expr :else Expr}
%%      '{:op :const :val Any} ...))
%% (ann tree-str [Expr -> String])
%% (defmulti tree-str (fn [e :- Expr] (get :op e)))
%% (defmethod tree-str :if [{:keys [test then else]}] 
%%   (str (tree-str test) (tree-str then) (tree-str else)))
%% (defmethod tree-str :const [{:keys [val]}] 
%%   (str val))
%% \end{minted}

%% \subsection{Contributions}

%% %The main idea is....... SPJ
%% In this paper we present a practical type system for Clojure.

%% \begin{itemize}

%%   \item We formalise Typed Clojure and prove that typed code
%%     \begin{itemize}
%%       \item disallows null-pointer exceptions, (Section ?)
%%       \item statically resolves Java interoperability (Section ?)
%%     \end{itemize}
%%   \item We reveal the surprising synthesis between Clojure multimethods
%%         and occurrence typing.
%%   \item We 
%% \end{itemize}

% what are the goals of Typed Clojure?
% where do they differ from Typed Racket? why?

% Things todo
% - DONE small calculus based on previous TR formalisation
%   that demos the ``flow'' environment & not/intersection type
% - discussion and examples for flow filters
% - discussion on the not/difference type, real code where it helps
% - higher-rank polymorphism, show what it enables (eg. monads)
%   - also show more complicated example, making an extensible
%     type for clojure.core/conj, and why it's not straightforward
% - discussion on limitations of local type inference
% - demonstration of how we check multimethods
% - dealing with array covariance
% - note on how we avoid null-pointer exceptions
% - discussion on why Typed Clojure only uses the ``static'' part
%   of Typed Racket
% - the equality filter problem (and draft solution?), a proposition
%   that knows about binding aliasing. Sounds straightforward?

>>>>

\begin{figure}
\inputminted[firstline=5]{clojure}{code/demo/src/demo/parent2.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

Figure~\ref{fig:ex1} presents a simple program demonstrating many
aspects of our system, from simple type annotations to explicit
handling of Java's \java{null} (written \clj{nil}) in interoperation, as well as an
extended form of occurrence typing and method resolution of
Java interoperability based on static type information.

The \clj{parent} function has the type 
$$
\clj{['{:file (U nil File)} -> (U nil Str)]}
$$
which means that it takes a hash table whose \clj{:file} key maps to either
\clj{nil} or a \clj{File}, and it produces either \clj{nil} or a
\clj{String}. The \clj{parent} function uses the \clj{:file} keyword
as an accessor to get the file, checks that it isn't \clj{nil}, and
then obtains the parent by making a Java method call.

@@

\subsection{Contributions}

These are our main contributions.

\begin{itemize}
  \item We motivate, describe and evaluate Typed Clojure, an optional type system for Clojure
        that understands existing Clojure idioms.
  \item We present the theory and implementation of a type system
        based on \emph{occurrence typing} that statically prevents null-pointer 
        exceptions in typed code in the presence of Java interoperability, 
        applicable to existing JVM languages.
  \item We prove type soundness for Typed Clojure.
\end{itemize}

### 2015-05-15 overview.tex
delete ns form. delete some type hints. use `new` over dot.

\inputminted[firstline=1]{racket}{code/tr/example1.rkt}

@@

\newpage
Here is the above program in Typed Clojure.
\begin{exmp}
\inputminted[firstline=1]{clojure}{code/demo/src/demo/eg1.clj}
\label{example:conditionalflow}
\end{exmp}

This is a regular Clojure file compiled with
the Clojure compiler, with the \clj{ns} form
declaring a
\emph{namespace}
for managing var and class imports.
Here \clj{:require} declares a runtime dependency 
on
\clj{clojure.core.typed},
Typed Clojure's core namespace, and
\clj{:refer}
brings a collection of vars into scope by name.
The \clj{:refer-clojure} \clj{:exclude} option unmaps core vars from the current 
namespace---here we unmap \clj{clojure.core/fn},
which creates a function from a parameter vector and a body expression,
and import a typed variant of \clj{fn}
that supports type annotations.

The typed \clj{fn} supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.
Typed Clojure provides a \clj{check-ns} function to type check the current namespace.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
any number of types.

@@

The \clj{do} form sequences two expressions returning the latter,
\clj{throw} corresponds to Java's \java{throw}
and \clj{(class. args*)} is the syntax for Java constructors---that 
is a class name with a dot suffix as the operator followed the arguments to the constructor.

In this example a \clj{throw} expression guards \clj{(inc x)}, the increment function for numbers,
from being evaluated if \clj{x} is \clj{nil}, preventing
a possible null-pointer exception.

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{See \url{https://github.com/typedclojure/examples}
  for full examples. From here we omit \clj{ns} forms.}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

>>>>

\inputminted[firstline=3]{racket}{code/tr/example1.rkt}

@@

Here is the above program in Typed Clojure.
\begin{exmp}
\inputminted[firstline=5]{clojure}{code/demo/src/demo/eg1.clj}
\label{example:conditionalflow}
\end{exmp}

The \clj{fn} macro (provided by core.typed) supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
As in Typed Racket, \clj{U} creates an \emph{untagged union} type, which can take
any number of types.

@@

The fully expanded increment function in~\egref{example:doexception}
guards its final call with a number check, preventing
a possible null-pointer exception.
Without this check, the type system would reject the program.

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{See \url{https://github.com/typedclojure/examples}
  for full examples.}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).
